---
title: "Seattle Weather Prediction"
author: "Chitti Babu Yelugubandi"
date: '`r Sys.Date()`'
output: 
  rmdformats::readthedown:
    css: gvsu_readthedown_theme.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Seattle Weather Prediction Using Regression Analysis

## Project Objective: 

The objective of this project, titled "Seattle Weather Prediction Using Regression Analysis," is to develop a predictive model to forecast the maximum daily temperature (`temp_max`) in Seattle using historical weather data. By analyzing features such as minimum temperature (`temp_min`), precipitation, wind speed (`wind`), and weather type (`weather`), my aim to create an accurate and reliable model. This involves data preparation, exploratory data analysis, and model evaluation using metrics like RMSE and MAPE. The goal is to use the model to predict future temperatures and support weather-related decision-making in Seattle.

## Importing the libraries

```{r}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(corrplot)
library(lubridate)
library(caTools)
library(caret)

```

## Loading the Data

```{r}
# Loading the data set
weather_data <- read.csv('D:/STA 631/seattle-weather.csv')

# Printing the first few rows of the dataset
head(weather_data)

```

## Data Dimensions and Summary Statistics


```{r}
# Data dimensions
dim(weather_data)

```

```{r}
# Summary statistics
summary_stat <- summary(weather_data)

summary_stat
```

## Data Types

```{r}
# Check data types
str(weather_data)

```


## Distribution of Numerical Columns

```{r}
precipitation_hist <- ggplot(weather_data, aes(x = precipitation)) + 
  geom_histogram(binwidth = 1, fill = "#AFEEEE", color = "black") +
  ggtitle("Distribution of Precipitation")

temp_max_hist <- ggplot(weather_data, aes(x = temp_max)) + 
  geom_histogram(binwidth = 1, fill = "#FFFFCC", color = "black") +
  ggtitle("Distribution of Maximum Temp")

temp_min_hist <- ggplot(weather_data, aes(x = temp_min)) + 
  geom_histogram(binwidth = 1, fill = "#99FF99", color = "black") +
  ggtitle("Distribution of Minimum Temp")

wind_hist <- ggplot(weather_data, aes(x = wind)) + 
  geom_histogram(binwidth = 0.5, fill = "#99CCFF", color = "black") +
  ggtitle("Distribution of Wind Speed")

grid.arrange(precipitation_hist, temp_max_hist, temp_min_hist, wind_hist, nrow = 2, ncol = 2)

```
## Categorical Variable Frequency Analysis

```{r}
table(weather_data$weather)
```


## Distribution of Categorical Columns


```{r}
weather_data$weather <- as.factor(weather_data$weather)

weather_plot <- ggplot(weather_data, aes(x = weather)) + 
  geom_bar(fill = "gray", color = "black", alpha = 0.7) +
  labs(title = "Weather Distribution", x = "Weather Type", y = "Frequency")

temp_max_plot <- ggplot(weather_data, aes(x = weather, y = temp_max, fill = weather)) + 
  geom_boxplot() +
  labs(title = "Maximum Temperature by Weather Type", x = "Weather Type", y = "Maximum Temperature") +
  theme(legend.position = "none")

temp_min_plot <- ggplot(weather_data, aes(x = weather, y = temp_min, fill = weather)) + 
  geom_boxplot() +
  labs(title = "Minimum Temperature by Weather Type", x = "Weather Type", y = "Minimum Temperature") +
  theme(legend.position = "none")

precipitation_plot <- ggplot(weather_data, aes(x = weather, y = precipitation, fill = weather)) + 
  geom_boxplot() +
  labs(title = "Precipitation by Weather Type", x = "Weather Type", y = "Precipitation") +
  theme(legend.position = "none")

grid.arrange(weather_plot, temp_max_plot, temp_min_plot, precipitation_plot, ncol = 2)
```


## Bivariate analysis

```{r}
temp_precip_scatt <- ggplot(weather_data, aes(x = precipitation, y = temp_max)) +
  geom_point(color = 'orange', alpha = 0.5) +
  labs(title = "Maximum Temp vs Precipitation", x = "Precipitation", y = "Maximum Temp")

temp_min_scatt <- ggplot(weather_data, aes(x = temp_max, y = temp_min)) +
  geom_point(color = 'blue', alpha = 0.5) +
  ggtitle("Max Temp vs Min Temp") +
  labs(x = "Max Temp", y = "Min Temp")

precipitation_scatt <- ggplot(weather_data, aes(x = temp_max, y = precipitation)) +
  geom_point(color = 'green', alpha = 0.5) +
  ggtitle("Max Temp vs Precipitation") +
  labs(x = "Max Temp", y = "Precipitation")

wind_scatt <- ggplot(weather_data, aes(x = temp_max, y = wind)) +
  geom_point(color = 'red', alpha = 0.5) +
  ggtitle("Max Temp vs Wind Speed") +
  labs(x = "Max Temp", y = "Wind Speed")

grid.arrange(temp_precip_scatt, temp_min_scatt, precipitation_scatt, wind_scatt, nrow =2, ncol = 2)

```


## Multi-Variate analysis

```{r}
# Convert necessary columns to appropriate types
weather_data$weather <- as.factor(weather_data$weather)

# Multivariate Scatter Plot: Temp Max vs. Temp Min colored by Weather Type
multivariate_weather_plot <- ggplot(weather_data, aes(x = temp_min, y = temp_max, color = weather)) +
  geom_point(alpha = 0.7) +
  labs(title = "Temp Max vs Temp Min by Weather Type",
       x = "Min Temperature",
       y = "Max Temperature") +
  theme_minimal()

# Display the plot
print(multivariate_weather_plot)
```

## Correlation matrix
```{r}
# correlation matrix
cor_matrix <- cor(weather_data %>% select_if(is.numeric))

# Print the correlation matrix
print(cor_matrix)

# Visualize the correlation matrix
library(corrplot)
corrplot(cor_matrix, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45, addCoef.col = "black", 
         diag = FALSE, order = "hclust")

```

## Data Cleaning

```{r}
# Checking missing values

colSums(is.na(weather_data))
```
## Converting categorical columns to numerical columns

```{r}

# Label encoding for 'weather' column
weather_data$weather <- as.numeric(factor(weather_data$weather, levels = unique(weather_data$weather)))

# Scaling numerical columns
numerical_columns <- c("temp_max", "temp_min", "precipitation", "wind")
weather_data[numerical_columns] <- scale(weather_data[numerical_columns])

```
## Data standardization 
```{r}
# Standardization for numerical columns
weather_data$temp_max <- scale(weather_data$temp_max)
weather_data$temp_min <- scale(weather_data$temp_min)
weather_data$precipitation <- scale(weather_data$precipitation)
weather_data$wind <- scale(weather_data$wind)

```

## Data splitting
```{r}
# Splitting the data into training and testing sets (70% train, 30% test)
set.seed(123) 
split <- sample.split(weather_data$temp_max, SplitRatio = 0.7)
train_data <- subset(weather_data, split == TRUE)
test_data <- subset(weather_data, split == FALSE)

```

## Model Selection

```{r}
# Load necessary library
library(caret)

# Define training control for cross-validation
cv_control <- trainControl(method = "cv", number = 5)  # 5-fold cross-validation

# Train the base model using all predictors
base_model <- train(temp_max ~ temp_min + precipitation + wind + weather, 
                    data = train_data, method = "lm", trControl = cv_control)

# Train a reduced model using selected predictors based on domain knowledge
reduced_model <- train(temp_max ~ temp_min + precipitation, 
                       data = train_data, method = "lm", trControl = cv_control)

# Compare models using resampling
model_results <- resamples(list(base = base_model, reduced = reduced_model))

# Summary of the results
print(summary(model_results))

```
 
## Fitting Linear regression model 

```{r}

# Fitting a linear regression model to predict temp_max using all predictors
lm_model_all <- lm(temp_max ~ temp_min + precipitation + wind + weather, data = train_data)

# Printing summary of the linear regression model
summary(lm_model_all)

```

## Distribution of Residual analysis

```{r}
# Distribution of residuals
resid_hist <- ggplot(lm_model_all, aes(x = .resid)) + 
  geom_histogram(binwidth = 0.1, fill = "gray") +
  labs(title = "Histogram of Residuals", x = "Residuals", y = "Count")

# Variance of residuals
resid_fitted <- ggplot(lm_model_all, aes(x = .fitted, y = .resid)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Residuals vs. Fitted Values", x = "Fitted Values", y = "Residuals")

# For 2 plots
grid.arrange(resid_hist, resid_fitted, nrow = 2, ncol = 1, top = "Regression Assumptions")
```


## Predictions on test data
```{r}
# Predicting temp_max on the test set
predictions <- predict(lm_model_all, newdata = test_data)

```

## RMSE Calculation

```{r}
# Calculating RMSE
rmse <- sqrt(mean((test_data$temp_max - predictions)^2))
print(paste("Root Mean Squared Error (RMSE):", round(rmse, 2)))

```

## MAPE Calculation

```{r}
# Calculating MAPE
mape <- mean(abs((test_data$temp_max - predictions) / test_data$temp_max)) * 100
print(paste("Mean Absolute Percentage Error (MAPE):", round(mape, 2)))

```

